---
title: "Gov 50 Final Project"
author: "Kelly Olmos"
description: "My Final Project"
output:
  distill::distill_article:
    self_contained: false
---
Introduction

The research question I have set out to answer is whether or not the stance a Republican candidates takes on the results of the 2020 presidential election are related to their electoral success in the 2022 general elections. In other words, is there a statistically significant difference in the proportion of elected Republican candidates based on whether they denied or accepted the results of the 2020 presidential election. My hypothesis is that there is a statistically significant difference in the proportion of Republican candidates that are elected based on their stance on the results of the 2020 presidential election. After the 2022 general election, numerous news articles and political pundits talked about the role of denying or accepting the results of the 2020 presidential election on electoral success of a candidate. This question is interesting because the answer to this question could shed light on potential warnings for democracy. If denying the results of the 2020 presidential election turns out to be an influencing factor in electoral success it could encourage future candidats to reject results of fair elections, thereby threatening democracy as it increases misinformation and polarization. 


Data Section: 

I obtained most of the data in my imported data set from FiveThirtyEight's Github repository (https://github.com/fivethirtyeight/data/tree/master/election-deniers). The key dependent and independent variables I will look at are `Won or Lost their Race` and `Stance`. The independent variable is the stance a Republican candidate takes on the 2020 presidential election results. Those stances include “Fully denied”, “Raised questions”, “Accepted with reservations”, “Fully accepted”, “Avoided answering”, and “No comment” which are all self-explanatory. FiveThirtyEight includes the sources for the stances of candidates in the `Source` variable. This variable is measured through the public responses of these candidates, their actions, and their lack of response to FiveThirtyEight's survey on their stances. 

 (2) a data section that briefly describes the data source, describes how the key dependent and independent variables are measured (e.g., a survey, statistical model, or expert coding), and also produces a plot that summarizes the dependent variable; (3) a results section that contains a scatterplot, barplot, or boxplot of the main relationship of interest and output for the main regression of interest; and (4) a brief (one paragraph) concluding section that summarizes your results, assesses the extent to which you find support for your hypothesis, describes limitations of your analysis and threats to inference, and states how your analysis could be improved (e.g., improved data that would be useful to collect).

For the data section, you should note if your research design is cross-sectional (most projects will be of this type) or one of the other designs we discussed (randomized experiment, before-and-after, differences-in-differences). For the results section, you should interpret (in plain English) the main coefficient of interest in your regression. You should also comment on the statistical significance of the estimated coefficient and whether or not you believe the coefficient to represent a causal effect.

Here is a rubric for the the core components of the final project:

Data section: 2-3 paragraphs + plot visualizing main outcome of interest. (3pts)
Results section: plot of main analysis + regression output + 2-3 paragraphs of description and interpretation of the plots and regression (including interpreting the main coefficient of interest and describing if it is statistically significant and if we should interpret it causally). This section could be longer if you choose to include additional analyses. (8pts)
Conclusion section: 1 paragraph (i) summarizing results and assessing the extent to which you find support for your hypothesis; (ii) describing limitations of the analysis and threats to inference (missing data, confounding, etc), and stating how you could improve your analysis if you had more time/money. (2pts)
To earn full credit on the visualizations and regression output, they should use informative labels and names and have a small number of digits presented. Tables should use kable or modelsummary to format output nicely. Generally speaking, the final report should mostly be readable by a person who hasn’t taken Gov 50.

## Project thoughts
In Republican-leaning states, Republican candidates who deny the 2020 election results have greater electoral success and the opposite occurs in democratic-leaning states. 
[concede that this system is not perfect bc it is the aggregate of the state and we know that states can be politically fragmented with some urban centers being much more liberal than rural place maybe even use IL as an example]

Describe your explanatory variable(s) of interest and how it is measured. Importantly, we need to observe variation in this variable in order to study it!

The explanatory variable of interest is the Republican candidate's stance on the 2020 election results. It is a categorical variable measured 
"Candidates in the “Fully denied” category have explicitly said the 2020 election was illegitimate and/or took legal measures to try and overturn the election. Examples of the latter include members of Congress who voted against certifying election results or politicians who signed an amicus brief for a Texas lawsuit that challenged election results in four states President Biden won.
Candidates in the “Raised questions” category raised questions or concerns about the election but haven’t outright denied or affirmed it.
Candidates in the “Accepted with reservations” category have accepted Biden’s victory but have still raised questions or concerns about fraud.
Candidates in the “Fully accepted” category have accepted Biden’s victory without reservations.
Candidates in the “Avoided answering” category refused or avoided directly answering questions about the election (e.g., by changing the subject when asked about it).
Candidates in the “No comment” category have not made a public comment on the election (as far as we could tell) and did not respond to our inquiries on the topic." 


Describe your outcome variable of interest and how it is measured.
The outcome variable of interest is whether or not the Republican candidate won their race. 

What observed pattern in the data would provide support for your hypothesis? More importantly, what observed pattern would disprove your hypothesis?


Introduction: describe the research question and main hypothesis; describe why it is important. (1-2 paragraphs) (2pts)
Data section: 2-3 paragraphs + plot visualizing main outcome of interest. (3pts)

I want to investigate the relationship between the stance a Republican candidate takes on the results of the 2020 election and their electoral success in the 2022 general elections? 
How does the partisan lean of the state affect the answer to this question? 

Results section: plot of main analysis + regression output + 2-3 paragraphs of description and interpretation of the plots and regression (including interpreting the main coefficient of interest and describing if it is statistically significant and if we should interpret it causally). This section could be longer if you choose to include additional analyses. (8pts)
Conclusion section: 1 paragraph (i) summarizing results and assessing the extent to which you find support for your hypothesis; (ii) describing limitations of the analysis and threats to inference (missing data, confounding, etc), and stating how you could improve your analysis if you had more time/money. (2pts)

[finish later]

```{r Loading Packages and Data Source(CSV File)}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(infer)
library(pscl)
election_deniers <- read_csv("~/Github Gov 50/FinalProjectGOV50/Gov50FinalProjectElectionDeniers/fivethirtyeight_election_deniers_Gov50_final.csv")
```
```{r Making the Data Set Neater by Selecting and Renaming Variables}
election_data <- election_deniers |>
  select(Stance, `Dem or Rep State`, `Won or Lost their Race`, `The FiveThirtyEight partisan lean of every state`) 

election_data <- 
  rename(election_data, President = `Dem or Rep State`) |>
  rename(Elected_or_not = `Won or Lost their Race`) |>
  rename(State_partisan_lean = `The FiveThirtyEight partisan lean of every state`)

election_data
```

I ran a 2-sample permutation test to compare the proportion of Republican candidates in the 2022 general elections who denied versus accepted the results of the 2020 election and their electoral success in the 2022 general election. For the purpose of simplifying the categorical variable Stance, I made it a binary variable where if candidates fully denied they were categorized as having denied the results of the 2020 election and every other option was considered accepting because these candidates did not take action to mark the election as illegitimate even if they might have raised concerns, they overall accepted the results. My two different samples are Republican candidates who denied the results of the 2020 election and Republican candidates who accepted the results of the 2020 election. The proportions I'm comparing are the proportion of candidates who won their race. The null hypothesis is that there is no difference in the proportion of Republican candidates who are elected based on their stance on the 2020 election. The permutation test provided a p-value of approximately 0, therefore, at alpha = 0.05 level, the difference in proportion of elected Republican candidates is different. I can reject the null hypothesis that there is no difference in the proportion of Republican candidates who were elected based on whether or not they denied the results of the 2020 election.

```{r Hypothesis Test (2 Sample Permutation Test with Infer)}
election_prop <- election_data |>
  mutate(partisan_leaning = if_else(State_partisan_lean >= 0, "Conservative Leaning", "Liberal Leaning"),
         elected = if_else(Elected_or_not == "Elected", 1, 0),
         stance = if_else(Stance == "Fully denied", "Denied", "Accepted")) |>
  group_by(stance) |>
  summarise(elected_prop = mean(elected))

knitr::kable(election_prop, col.names = c("Stance", "Prop. of Rep. Candidates Elected"))

election_data_modified <- election_data |>
  mutate(partisan_leaning = if_else(State_partisan_lean >= 0, "Conservative Leaning", "Liberal Leaning"),
         elected = if_else(Elected_or_not == "Elected", "1", "0"),
         stance = if_else(Stance == "Fully denied", "Denied", "Accepted")) |>
  specify(elected ~ stance, success = "1") |>
  calculate(stat = "diff in props", order = c("Denied", "Accepted")) 

knitr::kable(election_data_modified, col.names = c("Difference in Proportion of Electeds by Stance (Denied prop. - Accepted Prop.)"))

set_hypothesis <- election_data |>
  mutate(elected = if_else(Elected_or_not == "Elected", "1", "0"),
         stance = if_else(Stance == "Fully denied", "Denied", "Accepted")) |>
  specify(elected ~ stance, success = "1") |> 
  hypothesize(null = "independence") 
  

null_dist <- set_hypothesis |>
  specify(elected ~ stance, success = "1") |> 
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in props", order = c("Denied", "Accepted"))

null_dist |>
  visualize() +
  labs(
    x = "Diff. in Prop.",
    y = "Count",
    title = "Simulation-Based Null Distribution of Diff. in Prop."
  )

pval <- null_dist |>
  get_p_value(obs_stat = election_data_modified, direction = "both")

knitr::kable(pval, col.names = "P-value")
```

```{r Graph Visualizations}


# Create a boxplot using ggplot and dplyr
election_data |>
  mutate(stance = if_else(Stance == "Fully denied", "Denied", "Accepted")) |>
  filter(stance == "Denied") |>
  ggplot(aes(x = Elected_or_not, y = State_partisan_lean, fill = Elected_or_not)) +
  geom_boxplot() +
  labs(title = "Elecion Deniers: State Partisan Lean vs Election Outcome", x = "Elected or not", y = "State Partisan Lean")


contingency_table <- table(election_data$Stance, election_data$Elected_or_not)

knitr::kable(contingency_table)

mosaic_chart <- election_data |>
  mutate(stance = if_else(Stance == "Fully denied", "Denied", "Accepted")) |>
  ggplot(aes(x = Elected_or_not, fill = stance)) +
  geom_bar(position = "fill") +
  coord_flip() +
  labs(title = "Electoral Success by Stance", 
       x = "Electoral Success",
       y = "Proportion", 
       fill = "Stance")
mosaic_chart
```
```{r}
regression <- election_data |>
  mutate(elected = if_else(Elected_or_not == "Elected", 1, 0),
         stance = if_else(Stance == "Fully denied", 0, 1),
         state = election_deniers$State,
         incumbency = election_deniers$Incumbent,
         office = election_deniers$Office,
         president = election_deniers$`Dem or Rep State`,
         )

fit <- glm(elected ~ stance + President + incumbency + state + office , regression, family = "binomial") 
s <- summary(fit)$coefficients
s
modelsummary::modelsummary(fit)
fit
library(pscl)
pscl::pR2(fit)["McFadden"]

ggplot(regression, mapping = aes(x = stance, y = elected))  + geom_smooth(method = "lm")

```





